Missing data: Current practice in football research and recommendations for improvement

### 2020-09-04 paper/paper.Rmd
changes in discussion.

Our survey of 136 articles published on football related topics found
that only about one in 20 papers reported missing data. It is reasonable
to assume that this result is representative of current practices in the
wider sports science literature. This finding may be largely explained
by a lack of awareness and education on missing data practices (Kristin
BJSM paper), similar to other areas of sciences (refs-education and
ecology examples)(2). Below we discuss several poor practices in
relation to missing data, and provide recommendations for exploring,
visualising and reporting missingness.

@@

Case study: *Unreported missing values*. While not always the case, even if
not reported, missing data can be obvious. A recent paper examined the
relationship between match performance indicators and outcome, in
Australian Football between the 2001 and 2016. Problematically, at least
one of the 91 team performance indicators included in the analysis was
not available over the entire 2001-2016 period [@Young2019]. The variable
'meters gained' was not collected over the specified duration [@ref].
This was not noted anywhere in the paper, or in the supplementary
materials. It is unknown whether imputation was undertaken. Without
these details, it is difficult for other analysts and researchers to use
and/or extend the ideas presented in the paper, or reproduce the
analysis. It is also difficult for a reader to evaluate how the
missingness could bias the results. Unfortunately, the absence of a
missing data statement when data are missing is not unique to the study
above [@Young2019]. Authors should include a missing data statement,
irrespective of whether data is missing, or not.

Case study: *Mean imputation.* The default of most statistical software for
handling missing data in modelling is listwise deletion. Listwise
deletion removes entire rows that contain missing observations from the
analysis. At best, listwise deletion reduces the power in analysis by
reducing the sample size. At worst, it introduces bias. For example, if
injured participants are removed from a study on injury prevention.
Imputing values removes the need to conduct listwise deletion. However,
it is critical how these values are imputed, and that the imputation
method is documented. One imputation method is imputing the mean value
(e.g., participant or group average) [@Young2019]. While imputing the mean
preserves the study sample size (and point estimates), it also reduces
the variance, can alter the relationship between the variable with
missing observations and other variables, and can bias (underestimate)
standard errors [@Scheffer2002].
Smaller standard errors typically reduce *p*-values, which may lead to
incorrect inference. Mean imputation should be avoided. An imputation
strategy that incorporates information from other related variables in
the data -- such as linear regression, or K nearest neighbours -- should
be considered. This is discussed in the recommendations section.

Case study: *Not evaluating the effect, or choice, of imputation.* The
imputation method used to impute missing data has the potential to
affect the outcome of a study. Therefore, it is important that authors
understand, and document, how imputation affected the goal of the
analysis (i.e., inference, prediction, or both). For example,
understanding the implications in using mean imputation, compared to
K-nearest neighbour imputation -- the effect on parameter estimates
('significant' versus 'not significant'), and the (un)certainty of the
coefficients (i.e., less/more). When using imputation, authors need to
perform a sensitivity analysis (ref circulation article-Nick). This is
discussed below.

>>>>

Our survey of 136 articles published on football related topics found
that only about one in 20 papers reported missing data. It is reasonable
to assume that this result is representative of current practices in the
wider sports science literature. This finding may be largely explained
by a lack of awareness and education on missing data practices (https://bjsm.bmj.com/content/early/2020/08/19/bjsports-2020-102607.abstract), 
similar to other areas of sciences (refs-education and
ecology examples)(2). Below we discuss several practices to avoid in
relation to missing data, and provide recommendations for exploring,
visualising and reporting missingness.

@@

*Unreported missing values*. While not always the case, even if
not reported, missing data can be obvious. A recent paper examined the
relationship between match performance indicators and outcome, in
Australian Football between the 2001 and 2016. Of the 91 team performance indicators included in the analysis,
at least one variable (i.e., 'meters gained') was not available over the entire 2001-2016 period [@Young2019]**Check this reference**.
This was not noted anywhere in the paper, or in the supplementary
materials. It is unknown whether imputation was undertaken. Without
these details, it is difficult for other analysts and researchers to use
and/or extend the ideas presented in the paper, or reproduce the
analysis. It is also difficult for a reader to evaluate how the
missingness could bias the results. The absence of a
missing data statement when data are missing is not unique to the study
above [@Young2019]**Check this reference**. Authors should include a missing data statement,
irrespective of whether there is data missing, or not.

*Mean imputation.* The default of most statistical software for
handling missing data in modelling is listwise deletion. Listwise
deletion removes entire rows that contain missing observations from the
analysis. At best, listwise deletion reduces the power in analysis by
reducing the sample size. At worst, it introduces bias. For example, if
injured participants are removed from a study on injury prevention.
Imputing values removes the need to conduct listwise deletion. However,
it is critical how these values are imputed, and that the imputation
method is documented. One imputation method is imputing the mean value
(e.g., participant or group average) [@Young2019]**Check this reference**. While imputing the mean
preserves the study sample size (and point estimates), it also reduces
the variance, can alter the relationship between the variable with
missing observations and other variables, and can bias (underestimate)
standard errors [@Scheffer2002].
Smaller standard errors typically reduce *p*-values, which may lead to
incorrect inference. Mean imputation should be avoided. An imputation
strategy that incorporates information from other related variables in
the data -- such as linear regression, or K nearest neighbours -- should
be considered. This is discussed in the recommendations section.

*Not evaluating the effect, or choice, of imputation.* The
imputation method used to impute missing data has the potential to
affect the outcome of a study. Therefore, it is important that authors
understand, and document, how imputation affected the goal of the
analysis (i.e., inference, prediction, or both). For example,
understanding the implications in using mean imputation, compared to
K-nearest neighbour imputation -- the effect on parameter estimates
('significant' versus 'not significant'), and the (un)certainty of the
coefficients (i.e., less/more). When using imputation, authors need to
perform a sensitivity analysis (ref circulation article-Nick). This is
discussed below.
